Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-05-12 paper/main.tex
first draft of RNAseq methods

Orthologs of highly conserved single-copy genes were identified in each genome using
BUSCO. The {\it viridiplantae} set was used for plants and the {\it metazoa} set for animals.
The training genbank files were generated directly by BUSCO, by utilizing the ``--long"  % inline code?
parameter. For plants, the entire retraining could be performed as above; however, for animals
we randomly selected only half BUSCO-generated training set, which resulted in a training set 
size and runtime comparable to the plants.

What Marvin did, why we don't use Genic F1 for comparison, etc

>>>>

Orthologs of highly conserved generally single-copy genes were identified in each genome using
BUSCO. The {\it viridiplantae} set was used for plants and the {\it metazoa} set for animals.
The training genbank files were generated directly by BUSCO, by utilizing the ``--long"  % inline code?
parameter. For plants, the entire retraining could be performed as above; however, for animals
we randomly selected only half the BUSCO-generated training set, which resulted in a training set 
size and runtime comparable to the plants (~400 genes and several days per species). For animals
the training was carried out with the subsetted training file and using the following scripts provided
by Augustus. An untrained model was setup with "new\_species.pl", and the model was fit by running
"etraining" before and after the major hyper-parameter optimization with "optimize\_augustus.pl".
Using the trained model for each species, we ran the main prediction ("augustus") with ``--UTR=off"
and ``--gff3=on". The gffs produced by Augustus were imported into GeenuFF and exported as h5
files in the same manor as the reference, allowing direct comparison with reference and 
Helixer predictions. While tenable for some 237 species, this method allowed neither training
nor prediction of UTR regions with Augustus, so the metric Sub Genic F1, which excludes UTR,
was used for all comparisons between Augustus and Helixer.

\subsection{Evaluation against independent RNAseq data}
\label{sec:rnaseq}
The reference annotations were created with existing tools and largely with a pipeline incorporating
{\it de novo} gene predictions with RNAseq and homology data. The references, like any
data, are expected to contain errors; but here we were particularly concerned because
the {\it de novo} gene caller used by the references was Augustus for many species (...)
% todo list and cite examples (above and below)
and a fundamentally similar HMM-based tool for many more (...), which would potentially bias 
reference based comparisons of Helixer's and Augustus' performance.

We therefore downloaded and processed public RNAseq data to obtain an independent option 
for evaluating model performance. We selected three each of plant and animal genomes for
detailed evaluation with RNAseq. These were selected to have relatively good 
({\it Manihot esculenta} and {\it Papio anubis}), 
typical ({\it Medicago truncatula} and {\it Equus caballus}), 
and poor ({\it Theobroma cacao} and {\it Petromyzon marinus})
performance compared to Augustus \ref{} % supp phylogenies maybe?
within our generalizable range (i.e. excluding algae and
invertebrates). Selections were further constrained by the availability of stranded RNAseq
data.

% todo, add versions to everything below
For each of these six species the following search was performed on Sequence Read Archive
'(("<species name>"[Organism] OR <species name>[All Fields]) AND stranded[All Fields]) 
AND ("biomol rna"[Properties] AND "library layout paired"[Properties])'. In so far as 
more than 50 samples were identified, every Nth sample was selected so that about 50
samples were chosen for further processing, as listed here \ref{}. % todo make a supp table
Each sample was quality controlled with FastQC \citep{}, before and after trimming
with Trimmomatic \citep{}, and then mapped to the genome with Hisat2 \citep{}. 
Samtools \citep{} was used for the sorting and indexing that was necessary both for
RNAseq mapped metric collection with PicardTools \citep{} and custom downstream 
processing below. MultiQC \citep{} was used to summarize the results. The above was 
automated and the results visualized with the code here (... make available and link to RNAseq stuff).
Samples were filtered to those that had {\it relatively} high mapping rates, particularly to
exonic regions; that had low 3' bias, normal FastQC and Trimmomatic stats, and that
were confirmed to be stranded with 2nd strand as sense strand. If more remained, seven
of the high quality samples were selected randomly.

Finally, the selected, quality, mapped RNAseq samples were quantified to
get the coverage (number of reads matching; i.e. cigar =, M or X),
and spliced coverage (number of reads with gap or splice; i.e. cigar N or D)
for every base pair in the genomes as implemented in the "training\_rnaseq.py"
script.
